\documentclass[journal]{IEEEtran}

% --- Packages ---
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{url}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, positioning, arrows.meta}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{fancyhdr}

% --- Custom Header and Footer ---
\fancypagestyle{firststyle}{
    \fancyhf{}
    \renewcommand{\headrulewidth}{0pt}
    \fancyhead[L]{\small\bfseries 2026 ICS UNDERGRADUATE RESEARCH SYMPOSIUM}
    \fancyfoot[C]{\small \copyright\ 2026 ICS University of the Philippines Los Ba\~nos}
}

% --- Title and Author Details ---
\title{Pinoy Speak: A Trend Analysis and Contextual Lexicon System for Informal Filipino Text}
\author{Carl Timothy E. Clemente and Concepcion L. Khan}

\begin{document}

\maketitle
\thispagestyle{firststyle}

\begin{abstract}
Informal Filipino and Taglish evolve rapidly in online spaces, making them difficult to process using static dictionaries and conventional Natural Language Processing (NLP) resources. This paper presents Pinoy Speak, a deployed full-stack system for detecting, organizing, and explaining emerging Filipino internet slang. The system collects public text from selected Filipino Reddit communities, YouTube comments, and user chat interactions. As of the current implementation, it has archived 175,481 records and maintains approximately 104 verified slang entries. Candidate terms are processed through a pipeline that combines calamanCy tokenization, subword novelty signals, FastText nearest-neighbor evidence, temporal burstiness analysis, morphological filtering, and context-window disambiguation for ambiguous slang words. Verified candidates are enriched through a separate LLM pipeline, while the user-facing ``Kuya Slang'' tutor uses retrieval-augmented access to the live lexicon. Evaluation consisted of a 79-entry lexicon validation with three ratings per entry and a 36-response consented beta usability survey. Lexicon validation produced high definition correctness ratings (mean = 4.83/5) and example naturalness ratings (mean = 4.75/5), while beta usability results showed strong user acceptance across the dictionary, tutor, analyzer, navigation, and trend modules.
\end{abstract}

\begin{IEEEkeywords}
Informal Filipino, Taglish, slang detection, lexical trends, natural language processing, FastText, calamanCy, retrieval-augmented generation.
\end{IEEEkeywords}

% --- Include Sections ---
\input{sections/1_introduction}
\input{sections/2_literature_review}
\input{sections/3_methodology}
\input{sections/4_results}
\input{sections/5_conclusion}

% --- References ---
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}